---
output: github_document
---

**rocTree**
---

[![Project Status: Active – The project has reached a stable, usable state and is being actively developed.](http://www.repostatus.org/badges/latest/active.svg)](http://www.repostatus.org/#active)
[![minimal R version](https://img.shields.io/badge/R%3E%3D-3.4.0-6666ff.svg)](https://cran.r-project.org/)
[![CRAN_Status_Badge](http://www.r-pkg.org/badges/version/rocTree)](https://cran.r-project.org/package=rocTree)
[![packageversion](https://img.shields.io/badge/Package%20version-1.1.0-orange.svg?style=flat-square)](commits/master)
[![Travis-CI Build Status](https://travis-ci.org/stc04003/rocTree.svg?branch=master)](https://travis-ci.org/stc04003/rocTree)
[![AppVeyor Build Status](https://ci.appveyor.com/api/projects/status/github/stc04003/rocTree?branch=master&svg=true)](https://ci.appveyor.com/project/stc04003/rocTree)
[![Last-changedate](https://img.shields.io/badge/last%20change-`r gsub('-', '--', Sys.Date())`-yellowgreen.svg)](/commits/master)

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

### ROC-guided survival trees and ensembles
***

### Development

The package is under active development.

### Installation

You can install `rocTree` from **GitHub** with:

```{r gh-installation, eval = FALSE}
## install.packages("devtools")
devtools::install_github("stc04003/rocTree")
```

### Description
The `rocTree` provides implementations to a unified framework for tree-structured analysis with censored survival outcomes.
Different from many existing tree building algorithms, the `rocTree` package incorporate time-dependent covariates by constructing 
a time-invariant partition scheme on the survivor population. The partition-based risk prediction function is constructed 
using an algorithm guided by the Receiver Operating Characteristic (ROC) curve. 
Specifically, the generalized time-dependent ROC curves for survival trees show that the target hazard function yields the highest ROC curve. 
The optimality of the target hazard function motivates us to use a weighted
average of the time-dependent area under the curve (AUC) on a set of time points to
evaluate the prediction performance of survival trees and to guide splitting and pruning.
Moreover, the `rocTree` package also offers a novel ensemble algorithm, where the ensemble is on unbiased
martingale estimating equations. 

### Online documentations
[Online document](https://www.sychiou.com/rocTree/index.html) includes:

* Package vignette on [simulating data used in examples](https://www.sychiou.com/rocTree/articles/rocTree-sim.html).
* Package vignette on [growing time-invariant survival trees](https://www.sychiou.com/rocTree/articles/rocTree-tree.html).
* Package vignette on [growing survival forests](https://www.sychiou.com/rocTree/articles/rocTree-forest.html).

## Reference

Yifei Sun, Sy Han Chiou, Mei-Cheng Wang. ROC-Guided Survival Trees and Ensembles. [arXiv:1809.05627](https://arxiv.org/abs/1809.05627).