---
title: "Generating simulated data in `rocTree` package"
author: Sy Han (Steven) Chiou
date: "`r Sys.Date()`"
output:
  rmarkdown::html_document:
    toc: true
    toc_depth: 2
    toc_float: true
bibliography: ../inst/bib/tree.bib
vignette: >
  %\VignetteIndexEntry{Fitting semiparametric regression model to recurrent event data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}{inputenc}
---

```{r default, include = FALSE}
knitr::opts_chunk$set(prompt = TRUE, comment = "", collapse = TRUE)
```	

In this vignette, we demonstrate how to use the `simu` function in __`rocTree`__ package
to generate simulated data from various scenarios.

## Notations \& Simulation Settings

Let $Z$ be a $p$-dimensional vector of possible time-dependent covariate and $\beta$ be the vector of regression coefficient. 
The function `simu` generates survival times ($T$) under the following scenarios:

#### **Time-independent covariate**

**Scenario 1.1, proportional hazards model:**

Survival times are generated from the hazard function
$$\lambda(t|Z) = \lambda_0(t)\exp\{-0.5Z_1 + 0.5Z_2 - 0.5Z_3 + \ldots + 0.5Z_{10}\},$$
with $\lambda_0(t)=2t$.

**Scenario 1.2, proportional hazards model with noise variable:**

Survival times are generated from the hazard function
$$\lambda(t|Z) = \lambda_0(t)\exp\{2Z_1 + 2Z_2 + 0\cdot Z_3 + 0\cdot Z_4 + \ldots + 0\cdot Z_{10}\},$$
with $\lambda_0(t)=2t$.

**Scenario 1.3, proportional hazards model with nonlinear covariate effects:**

Survival times are generated from the hazard function
$$\lambda(t|Z) = \lambda_0(t) \exp\{2\sin(2\pi Z_1) + 2 |Z_2 - 0.5|\}, $$
with $\lambda_0(t)=2t$.

**Scenario 1.4, accelerated failure time model:**

Survival times are generated from 
$$\log(T) = -2 + 2Z_1 + 2Z_2 + \epsilon, $$
where $\epsilon\sim\mbox{N}(0, 0.5^2)$.

**Scenario 1.5, generalized gamma family:**

Survival times are generated from 
$$T = e^{\sigma\omega}, $$
where $\omega = \log(Q^2g)/Q$, $g$ follows gamma$(Q^{-2}, 1)$, $\sigma = 2Z_1$, $Q=2Z_2$.


#### **Time-dependent covariate**

**Scenario 2.1, dichotomous time dependent covariate with at most one change in value:**

Survival times are generated from the hazard function
$$\lambda(t|Z(t)) = e^{2Z_1(t) + 2Z_2}, $$
where $Z_1(t) = \theta I(t\ge U_0) + (1 - \theta)I(t<U_0)$, $\theta$ is a Bernoulli variable with equal probability, and
$U_0$ follows a uniform over $[0,1]$.

**Scenario 2.2, dichotomous time dependent covariate with multiple jumps:**

Survival times are generated from the hazard function
$$\lambda(t|Z(t)) = e^{2Z_1(t) + 2Z_2}, $$
where $Z_1(t) = \theta\left[I(U_1 \le t < U_2) + I(U_3\le t)\right] + (1 - \theta)\left[I(t < U_1) + I(U_2\le t < U_3)\right]$,
$\theta$ is a Bernoulli variable with equal probability and $U_1\le U_2\le U_3$ are the first three terms of a stationary Poisson process with rate 10.

**Scenario 2.3, proportional hazard model with a continuous time dependent covariate:** 

Survival times are generated from the hazard function 
$$\lambda(t|Z(t)) = 0.1 e^{Z_1(t) + Z_2}, $$
where $Z_1(t)=kt+b$, $k$ and $b$ are independent uniform random variables over $[1, 2]$.

**Scenario 2.4, non-proportional hazard model with a continuous time dependent covariate:** 

Survival times are generated from the hazard function 
$$\lambda(t|Z(t)) = 0.1 \cdot\left[1 + \sin\{Z_1(t) + Z_2\}\right],$$
where $Z_1(t)=kt+b$, $k$ and $b$ are independent uniform random variables over $[1, 2]$.

**Scenario 2.5, non-proportional hazard model with a nonlinear time dependent covariate:** 

Survival times are generated from the hazard function 
$$\lambda(t|Z(t)) = 0.1 \cdot\left[1 + \sin\{Z_1(t) + Z_2\}\right],$$
where $Z_1(t)=2kt\cdot\{I(t>5) - 1\}$, $k$ and $b$ are independent uniform random variables over $[1, 2]$.


## The `simu` function

The `simu` function can be used to generate survival times from the above scenarios.
The complete list of arguments in `simu` are as follow:
```{r reReg-args}
library(rocTree)
args(simu)
```
The arguments are as follows

* `n` an integer value indicating the number of subjects.
* `cen` is a numeric value indicating the censoring percentage; three levels, 0\%, 25\%, 50\%, are allowed.
* `scenario` can be either a numeric value or a character sting. This indicates the simulation scenario noted above. 
* `summary` a logical value indicating whether a brief data summary will be printed.

The `simu` places the simulated data in a `tibble` environment with the columns:

* `id` is the subject id.
* `Time` is the observed follow-up time.
* `death` is the death indicator; `death = 1` if an event (death) occurs and `death = 0` if censored.
* `z1`--`z10` are the possible time-dependent covariate.
* `k`, `b`, `U` are the latent variables used to generate $Z_1(t)$ in Scenario 2.1 -- 2.5.

## Example 1

We first generate a small dataset with `n = 5`, 25\% censoring rate, under scenario 1.2.
```{r ex1}
set.seed(2019)
dat1 <- simu(n = 5, cen = 0.25, sce = 1.2, summary = TRUE)
dat1
class(dat1)
```
In this scenario, the covariate information was observed at `Time = 0.0931`, `0.146`, `0.340`, and `0.423` for subject \#1, who died (`death = 1`) at `Time = 0.423`.
Since the covariate are time-independent, its values is invariant to time. 

## Example 2

The following codes generate a small dataset with `n = 5`, 50\% censoring rate, under scenario 2.1.
```{r ex2}
set.seed(2019)
dat2 <- simu(n = 5, cen = 0.5, sce = 2.1, summary = TRUE)
dat2
```
In this scenario, the covariate information was observed at `Time = 0.00883` and `0.102` for subject \#1, who died (`death = 1`) at `Time = 0.102`.
Similarly, the covaraite information was observed at `Time = 0.00883`, `0.102`, `0.105`, and `0.137` for subject \#2, who was censored (`death = 0`) at `Time 0.137`. 
Moreover, `z1` is a time-dependent covariate and its value changed from 1 (at `Time = 0.00883`) to 0 ( at `Time` $\ge$ `0.102`) for subject \#2.