---
title: "Fitting ROC-guided survival tree in `rocTree` package"
author: Sy Han (Steven) Chiou
date: "`r Sys.Date()`"
output:
  rmarkdown::html_document:
    toc: true
    toc_depth: 2
    toc_float: true
bibliography: ../inst/bib/tree.bib
vignette: >
  %\VignetteIndexEntry{Fitting semiparametric regression model to recurrent event data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}{inputenc}
---

```{r default, include = FALSE}
knitr::opts_chunk$set(prompt = TRUE, comment = "", collapse = TRUE)
```	

In this vignette, we demonstrate how to use the `rocTree` function in __`rocTree`__ package
to fit the proposed Receiver Operating Characteristic (ROC) guided survival tree.

## Simulated data

We will demonstrate the usage of `rocTree` function with a simulated data prepared by the [`simu`](https://www.sychiou.com/rocTree/articles/rocTree-sim.html) function.
```{R load}
library(rocTree)
set.seed(2019)
dat <- simu(n = 100, cen = 0.25, sce = 2.1, summary = TRUE)
```

## The `rocTree` function 

The complete list of arguments in `rocTree` are as follow:
```{R rocTree-arg}
args(rocTree)
```

The arguments are as follows

* `formula` is a formula object, with the response on the left of a `~` operator, and the predictors on the right. The response must be a survival object returned by the function `Surv` from the `survival` package. 
* `data` is an optional data frame to interpret the variables occurring in the `formula`.
* `id` an optional vector used to identify the longitudinal observations of subject's id. The length of `id` should be the same as the total number of observations. If `id` is missing, then each row of `data` represents a distinct observation from subjects and all covariates are treated as a baseline covariate. 
* `subset` an optional vector specifying a subset of observations to be used in the fitting process.
* `splitBy` a character string specifying the splitting algorithm. The available options are `CON` and `dCON` corresponding to the splitting algorithm based on the total concordance measure or the difference in concordance measure, respectively. The default value is `dCON`.
* `control` a list of control parameters. 

## The `control` options

The argument `control` defaults to a list with the following values:

* `tau` is the maximum follow-up time; default value is the 90th percentile of the unique observed survival times. 
* `M` is the maximum node number allowed to be in the tree; the default value is `1000`.
* `hN` is the smoothing parameter used in the Kernel; the default value is `tau / 20`.
* `minsp` is the minimum number of failure required in a node after a split; the default value is `20`. 
* `minsp2` is the minimum number of failure required in a terminal node after a split; the default value is `5`.
* `disc` is a logical vector specifying whether the covariates in `formula` are discrete (`1`). The length of `disc` should be the same as the number of covariates in `formula`.
* `prune` a logical vector specifying whether to prune the survival tree. If `TRUE`, a cross-validation procedure will be performed to determine the optimal subtree; the default value is `TRUE`.
* `nflds` is the number of folds used in the cross-validation. This argument is only needed if `prune = TRUE`. The default value is `10`.
* `Trace` is a logical vector specifying whether to display the splitting path; the default value is `FALSE`.
* `parallel` is a logical vector specifying whether parallel computing will be applied in cross-validation; the default value is `FALSE`. 
* `parCluster` is an integer value specifying the number of CPU cores to be used when `prune = TRUE` and `parallel = TRUE`. The default value is half of the number of CPU cores detected. 

## Growing a survival tree

We first load the `survival` package to enable `Surv`.
The fully grown (unpruned) time-invariant survival tree can be constructed as follow:
```{R tree}
library(survival)
system.time(fit <- rocTree(Surv(Time, death) ~ z1 + z2, id = id, data = dat))
```
The function `rocTree` returns an object of S3 class -- `rocTree` with the following major components:

* `Frame` is a data frame describe the resulting tree. The columns are 
  * `nd` node number.
  * `terminal` describe the node characteristic; `0` if a node is internal, `1` if a node is splitable, and `2` if a node is a terminal node.
  * `u` is the proportion being split in $\tau_L$.
  * `u2` is the minimum proporiton in $\tau_L$ across time.
  * `p` indicate which covariate was split at the node. 
  * `cut` is the covariate value being split at the node.
* `dfFinal` is a data.frame consists of unsmoothed hazard estimates at the observed survival times for the terminal nodes.

The time-invariant survival tree can be printed directly or with the generic function `print`.
```{r print}
fit
```
The survival tree is printed in the structure similar to that in the `data.tree` package. Setting `dt = FALSE` in the generic function prints the tree in a structure similar to that of the `rpart` package.
```{r print2}
print(fit, dt = FALSE)
```

## Plotting the survival tree

The survival tree can also be plotted with the `GraphViz`/`DiagrammeR` engine via the generic function `plot`.
```{r plot, fig.align = "center", fig.width = 6}
plot(fit)
```
The `plot` feature also allows the following useful options adopted from the [`Graphviz`/`DiagrammeR`](http://rich-iannone.github.io/DiagrammeR/graphviz_and_mermaid.html) environment to be passed to `option`:

* `rankdir` is a character string specifying the direction of the tree flow. The available options are top-to-bottom (`TB`), bottom-to-top (`BT`), left-to-right (`LR`), and right-to-left (`RL`); the default value is `TB`.
* `shape` is a character string specifying the shape style. Some of the available options are `ellipse`, `oval`, `rectangle`, `square`, `egg`, `plaintext`, `diamond`, and `triangle`. The default value is `ellipse`.
* `nodeOnly` is a logical value indicating whether to display only the node number; the default value is `TRUE`.
* `savePlot` is a logical value indicating whether the plot will be saved (exported); the default value is `FALSE`.
* `file_name` is a character string specifying the name of the plot when `savePlot = TRUE`. The file name should include its extension. The default value is `pic.pdf`.
* `file_type` is a character string specifying the type of file to be exported. Options for graph files are: `png`, `pdf`, `svg`, and `ps`. The default value is `pdf`.  

The following codes illustrate some of the different options.
```{r plot2, fig.align = "center", fig.width = 6, fig.show="hold"}
plot(fit, control = list(rankdir = "LR", shape = "rect"))
plot(fit, control = list(shape = "egg", nodeOnly = TRUE))
```

## Pruning the survival tree

Pruning reduces the complexity of the final classifier, and hence improves predictive accuracy by the reduction of overfitting.
Setting `prune = TRUE` in the `control` list will prune the survival tree. In the following example, we used five-fold cross-validation 
to choose the tuning parameter in the concordance-complexity measure:
```{R tree2, tidy = TRUE}
set.seed(2019)
system.time(fit2 <- rocTree(Surv(Time, death) ~ z1 + z2, id = id, data = dat, 
                control = list(prune = TRUE, nfld = 5)))
fit2
```
The resulting tree is much smaller than the un-pruned tree in `fit`. 

## Hazard estimates

The time-invariant partition considered allows a sparse model and an easy interpretation
of the decision rule. At each fixed time $t$, the tree partitions the survivor population and predicts the instantaneous failure risk.
Thus the interpretation at a fixed time point is along the same line as classification and regression trees. 
Since the risk within each terminal node changes with time, it is essential to look at the hazard curves of each terminal
The smoothed hazard estimates at terminal nodes can be easily plotted with the function `plotTreeHaz`, as demonstrated below.
In the following example, subjects in node 2 ($Z_1(t) \le 0.56$) yield the highest hazard for $t < 0.7$.
```{R haz, fig.align = "center"}
plotTreeHaz(fit2)
```

## Prediction

Suppose we have a new data that is generated as below:
```{R newDat}
newdat <- dplyr::tibble(Time = sort(unique(dat$Time)), 
                        z1 = 1 * (Time < median(Time)), 
                        z2 = runif(1))
newdat
```
The predicted survival curve can be plotted with the following codes.
```{R pred}
pred <- predict(fit2, newdat)
pred
plot(pred)
```
