---
title: "Fitting ROC-guided survival tree in `rocTree` package"
author: Sy Han (Steven) Chiou
date: "`r Sys.Date()`"
output:
  rmarkdown::html_document:
    toc: true
    toc_depth: 2
    toc_float: true
bibliography: ../inst/bib/tree.bib
vignette: >
  %\VignetteIndexEntry{Fitting semiparametric regression model to recurrent event data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}{inputenc}
---

```{r default, include = FALSE}
knitr::opts_chunk$set(prompt = TRUE, comment = "", collapse = TRUE)
```	

In this vignette, we demonstrate how to use the `rocTree` function in __`rocTree`__ package
to fit the proposed Receiver Operating Characteristic (ROC) guided survival tree.

## Simulated data

We will demonstrate the usage of `rocTree` function with a simulated data prepared by the [`simu`](https://www.sychiou.com/rocTree/articles/rocTree-sim.html) function.
```{R load, message = FALSE}
library(rocTree)
set.seed(0)
dat <- simu(n = 100, cen = 0.25, sce = 2.3, summary = TRUE)
head(dat)
```

## The `rocTree` function 

The complete list of arguments in `rocTree` are as follow:
```{R rocTree-arg}
args(rocTree)
```

The arguments are as follows

* `formula` is a formula object, with the response on the left of a `~` operator, and the predictors on the right. The response must be a survival object returned by the function `Surv` from the `survival` package. 
* `data` is an optional data frame to interpret the variables occurring in the `formula`.
* `id` is an optional vector used to identify the longitudinal observations of subject's id. The length of `id` should be the same as the total number of observations. If `id` is missing, then each row of `data` represents a distinct observation from subjects and all covariates are treated as a baseline covariate. 
* `subset` is an optional vector specifying a subset of observations to be used in the fitting process.
* `ensemble` is an optional logical value. If `TRUE` (default), ensemble methods will be fitted. Otherwise, the survival tree will be fitted.
* `splitBy` is a character string specifying the splitting algorithm. The available options are `CON` and `dCON` corresponding to the splitting algorithm based on the total concordance measure or the difference in concordance measure, respectively. The default value is `dCON`.
* `control` is a list of control parameters. 

## The `control` options

The argument `control` defaults to a list with the following values:

* `tau` is the maximum follow-up time; default value is the 90th percentile of the unique observed survival times. 
* `maxNode` is the maximum node number allowed to be in the tree; the default value is `500`.
* `numFold` is the number of folds used in the cross-validation. When `numFold > 0`, the survival tree will be pruned; when `numFold = 0`, the unpruned survival tree will be presented. The default value is `10`.
* `h` is the smoothing parameter used in the Kernel; the default value is `tau / 20`.
* `minSplitTerm` is the minimum number of baseline observations in each terminal node; the default value is `15`.
* `minSplitNode` is the minimum number of baseline observations in each splitable node; the default value is `30`. 
* `disc` is a logical vector specifying whether the covariates in `formula` are discrete (`TRUE`) or continuous (`FALSE`). The length of `disc` should be the same as the number of covariates in `formula`. When not specified, the `rocTree()` function assumes continuous covariates for all.
* `K` is the number of time points on which the concordance measure is computed. A less refined time grids (smaller `K`) generally yields faster speed but a very small `K` is not recommanded. The default value is 20.

## Growing a survival tree

We first load the `survival` package to enable the `Surv()` function, which we will use to specify the survival response.
The fully grown (un-pruned) time-invariant survival tree can be constructed as follow:
```{R tree}
library(survival)
system.time(fit <- rocTree(Surv(Time, death) ~ z1 + z2, id = id, data = dat, 
                           ensemble = FALSE, control = list(numFold = 0)))
```
The function `rocTree` returns an object of S3 class with the following major components:

* `Frame` is a data frame describe the resulting tree. The columns are 
  * `p` indicate which covariate was split at the node. 
  * `left` indicate the row number for the left child (if the current node splittable).
  * `right` indicate the row number for the right child (if the current node splittable).
  * `cutVal` indicate the covariate value (after transformation) being split at the node.
  * `cutOrd` indicate the rank of the covariate value being split at the node.
  * `nd` indicates the node number in display.
  * `is.terminal` describe the node characteristic; `0` if a node is internal, `1` if a node is a terminal node.

The time-invariant survival tree can be printed directly or with the generic function `print`.
```{r print}
fit
```
The survival tree is printed in the structure similar to that in the `data.tree` package. 

## Plotting the survival tree

The survival tree can also be plotted with the `GraphViz`/`DiagrammeR` engine via the generic function `plot`.
```{r plot, fig.align = "center", fig.width = 6}
plot(fit)
```
The `plot` feature also allows the following arguments adopted from the [`Graphviz`/`DiagrammeR`](http://rich-iannone.github.io/DiagrammeR/graphviz_and_mermaid.html) environment to be passed to `option`:

* `type` is an optional character string specifying what to plot. The available options are `tree` for plotting survival tree (default), `survival` for plotting the estimated survival probabilities for the terminal nodes, and `hazard` for plotting the estimated hazard for the terminal nodes. The following options are available only when `type = tree`.
* `output` is a string specifying the output type. The possible values are `graph` and `visNetwork`; `graph` (the default) renders the graph using the `grViz` function, and `visNetwork` renders the graph using the `visnetwork` function.
* `digits` the number of digits to print.
* `rankdir` is a character string specifying the direction of the tree flow. The available options are top-to-bottom (`TB`), bottom-to-top (`BT`), left-to-right (`LR`), and right-to-left (`RL`); the default value is `TB`.
* `shape` is a character string specifying the shape style. Some of the available options are `ellipse`, `oval`, `rectangle`, `square`, `egg`, `plaintext`, `diamond`, and `triangle`. The default value is `ellipse`.
* `nodeOnly` is a logical value indicating whether to display only the node number; the default value is `TRUE`.
* `savePlot` is a logical value indicating whether the plot will be saved (exported); the default value is `FALSE`.
* `file_name` is a character string specifying the name of the plot when `savePlot = TRUE`. The file name should include its extension. The default value is `pic.pdf`.
* `file_type` is a character string specifying the type of file to be exported. Options for graph files are: `png`, `pdf`, `svg`, and `ps`. The default value is `pdf`.  

The following codes illustrate some of the different options.
```{r plot2, fig.align = "center", fig.width = 6, fig.show="hold"}
plot(fit, rankdir = "LR", shape = "rect", digits = 2)
plot(fit, shape = "egg", nodeOnly = TRUE)
plot(fit, output = "visNetwork", digits = 2)
```

## Pruning the survival tree

Pruning reduces the complexity of the final classifier, and hence improves predictive accuracy by the reduction of overfitting.
Setting `prune = TRUE` in the `control` list will prune the survival tree. In the following example, we used five-fold cross-validation 
to choose the tuning parameter in the concordance-complexity measure:
```{R tree2, tidy = TRUE}
system.time(fit2 <- rocTree(Surv(Time, death) ~ z1 + z2, id = id, data = dat, 
                            ensemble = FALSE, control = list(numFold = 10)))
fit2
plot(fit2)
```
The resulting tree is much smaller than the un-pruned tree in `fit`. 

## Survival/Hazard estimates at terminal notes

The time-invariant partition considered allows a sparse model and an easy interpretation
of the decision rule. At each fixed time $t$, the tree partitions the survivor population and predicts the instantaneous failure risk.
Thus the interpretation at a fixed time point is along the same line as classification and regression trees. 
Since the risk within each terminal node changes with time, it is essential to look at the hazard curves of each terminal
The smoothed hazard estimates at terminal nodes can be easily plotted with the generic function `plot` with `type = "survival"` or `type = "hazard"`.
The feature is demonstrated below.
```{R haz, fig.align = "center"}
plot(fit2, type = "hazard")
```

## Prediction

Suppose we have a new data that is generated as below:
```{R newDat}
newdat <- dplyr::tibble(Time = sort(unique(dat$Time)), 
                        z1 = 1 * (Time < median(Time)), 
                        z2 = 0.5)
newdat
```
The predicted survival curve can be plotted with the following codes.
```{R pred}
(pred <- predict(fit2, newdat, type = "survival"))
plot(pred)
(pred <- predict(fit2, newdat, type = "hazard"))
plot(pred)
```
